%!TEX root = Thermodynamics.tex
\chapter{Introduction and Basic Concepts}
The purpose of this chapter is to identify and form a basic understanding of Thermodynamics jargon. This is the key to understanding any scientific area. We also will discuss the general technique used to solve Thermodynamics problems.
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Thermodynamics and Energy}
Thermodynamics is the study of \textbf{\textit{energy}}. A general definition for \textbf{\textit{energy}} is the ability to cause change in a \textbf{\textit{system}}. Though the name \textit{Thermodynamics}} suggests it is only concerned with studying \textbf{\textit{heat}}, today it is used to study \textbf{\textit{refrigeration}}, \textbf{\textit{power generation}}, and \textbf{\textit{matter}} properties.

\vspace{0.1in}
The \textbf{\textit{Conservation of Energy Principle}} is paramount to the study of Thermodynamics.

\vspace{0.1in}
\noindent \textbullet \hphantom{-} \textbf{\textit{Conservation of Energy Principle}}: \textbf{\textit{Energy}} cannot be created or destroyed, but only change from one form to another.

\vspace{0.1in}
The two main types of \textbf{\textit{energy}} are \textbf{\textit{Potential Energy}} and \textbf{\textit{Kinetic Energy}}. The change in \textbf{\textit{energy}} of a given system can be represented by:
%
\begin{equation}
\Delta E = E_{in} - E_{out}
\end{equation}
%
The study of Thermodynamics did not emerge until the first steam engines were constructed in 16978 and 1712 in England. Much like Sir Isaac Newton's Laws of Motion, Thermodynamics has laws that govern the study of energy:
%
\begin{enumerate}
\label{Laws of Thermodynamics}
\item \textbf{\textit{Energy}} is a thermodynamic property that cannot be created or destroyed, but can be converted from one form to another.\\
\item \textbf{\textit{Energy}} has \textit{quality} and \textit{quantity}.  \textbf{\textit{Processes}} occur in the direction of decreasing \textit{quality} of energy.
\end{enumerate}
%
These two laws were founded through the work of Rankine, Clausius, and Lord Kelvin.

\vspace{0.1in}
\textbf{\textit{Classical Thermodynamics}} uses a macroscopic approach to solving engineering problems, meaning that one does not need to know the behavior of the molecules of the system. \textbf{\textit{Statistical Thermodynamics}} studies the average behavior of large quantities of molecules on a microscopic scale. This text is mostly concerned with \textbf{\textit{Classical Thermodynamics}}.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Importance of Dimensions and Units}
Any physical object can be characterized by \textbf{\textit{dimensions}}. The magnitudes of the \textbf{\textit{dimensions}} are captured by \textbf{\textit{units}}. The \textbf{\textit{Primary Dimensions}} and corresponding SI units are listed below. All other units, the \textbf{\textit{Derived Units}} can be found using a combination of these measurements.
%
\begin{table}[b]
\begin{center}
\label{Primary Dimensions}
\begin{tabular}{lr}
\textbf{Dimension} & \textbf{Units}\\
\hline\\
Length & meter (m)\\
Mass & kilogram (kg)\\
Time & second (s)\\
Temperature & kelvin (K)\\
Electric Current & ampere (A)\\
Amount of Light & candela (cd)\\
Amount of Matter & mole (mol)
\end{tabular}
\end{center}
\end{table}
%
The majority of this reference will refer to quantities using SI units.

\newpage
Below are some conversions between important SI and US Customary Units:
%
\begin{equation}
1\text{ pound-mass (lbm)} = 0.45359\text{ kg}
\end{equation}
%
\begin{equation}
1\text{ foot (ft)} = 0.3048\text{ m}
\end{equation}
%
The SI \textbf{\textit{derived unit}} for \textbf{\textit{force}} is the \textbf{\textit{Newton}}. Newton's Second Law of Motion is given by:
%
\begin{equation}
\text{Force} = \text{Mass}\cdot\text{Acceleration}
\end{equation}
%
\begin{equation}
F = ma
\end{equation}
%
The \textbf{\textit{Newton}} is defined as the \textbf{\textit{force}} required to move a 1-kg object at a rate of 1 $\frac{\text{m}}{\text{s}^2}$. The US Customary equivalent is the \textbf{\textit{pound-force}}, which is the \textbf{\textit{force}} required to move 1 \textbf{\textit{slug}} (32.174 lbm) at a rate of 1 $\frac{\text{ft}}{\text{s}^2}$:
%
\begin{equation}
1\text{ Newton (N)} = 1 \hphantom{-}\frac{\text{kg}\cdot\text{m}}{\text{s}^2}
\end{equation}
%
\begin{equation}
1\text{ lbf} = 32.174 \hphantom{-} \frac{\text{lbm}\cdot\text{ft}}{\text{s}^2}
\end{equation}
%

Note: \textbf{\textit{weight}} is a \textbf{\textit{force}}, not a measurement of mass. A gravitational force exists between all objects; the magnitude of this force depends on the size of the objects and the distance between them. For most practical purposes, the gravitational constant, $g$, can be approximated as 9.807 $\frac{\text{m}}{\text{s}^2}.

\vspace{0.1in}
\textbf{\textit{Work}} is a form of \textbf{\textit{energy}}, and is defined as:
%
\begin{equation}
\text{Work} = \text{Force}\cdot\text{Distance \hphantom{-}(N}\cdot\text{m)}
\end{equation}
%
The standard SI \textbf{\textit{derived unit}} for \textbf{\textit{work}} is the \textbf{\textit{Joule}}:
%
\begin{equation}
1 \text{ Joule (J)} = 1 \text{ N}\cdot\text{m}
\end{equation}
%
The US Customary equivalent to the \textbf{\textit{Joule}} is the \textbf{\textit{British Thermal Unit}} (Btu):
%
\begin{equation}
1 \text{ Btu} = 1055.1 \text{ J}
\end{equation}
%
Another common unit for \textbf{\textit{energy}} is the \textbf{\textit{calorie}} (cal):
%
\begin{equation}
1 \text{ cal} = 4.1868 \text{ J}
\end{equation}
%
Note: when \textbf{\textit{calorie}} begins with a capital ``C," as in \textbf{\textit{Calorie}}, this is a \textit{kilo-calorie}:
%
\begin{equation}
1 \text{ Cal} = 1000 \text{ cal}
\end{equation}
%
Another \textbf{\textit{derived unit}} is used when adding a temporal component to \textbf{\textit{energy}}. The \textbf{\textit{Watt}} (W) is the standard SI unit to define \textbf{\textit{energy rates}}, or \textbf{\textit{power}}:
%
\begin{equation}
1 \text{ W} = 1 \hphantom{-} \frac{J}{s}
\end{equation}
%
The US Customary equivalent to \textbf{\textit{Watts}} is the \textbf{\textit{horsepower}} (hp):
%
\begin{equation}
1 \text{ hp} = 756 \text{ W}
\end{equation}
%
These conversions are important to make calculations with \textit{dimensional homogeneity}. Not paying attention to units is a common mistake for errors to propogate throughout a problem.
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Systems and Control Volumes}
A \textbf{\textit{system}} is defined as a quantity of \textbf{\textit{matter}} or a region in space chosen for study. Anything that is not part of the \textbf{\textit{system}} is referred to as the \textbf{\textit{surroundings}}. A hypothetical barrier can be used to separate the \textbf{\textit{system}} from the \textbf{\textit{surroundings}}.

\vspace{0.1in}
\textbf{\textit{Systems}} can be defined as either \textit{open} or \textit{closed}. A \textbf{\textit{Closed System}} or \textbf{\textit{control mass}} has a constant amount of \textbf{\textit{mass}} in the \textbf{\textit{system}} at all times; thus, no \textbf{\textit{mass}} can enter or leave the \textbf{\textit{system}} by crossing the barrier that separates it from the \textbf{\textit{surroundings}}. However, \textbf{\textit{energy}} can cross the barrier, and the volume of a \textbf{\textit{closed system}} does not need to remain fixed. A common example of a \textbf{\textit{closed system}} is a piston cylinder. A special case of a \textbf{\textit{closed system}} when energy is also not allowed to cross the barrier is an \textit{isolated system}}.

\vspace{0.1in}
\textbf{\textit{Open Systems}} or \textbf{\textit{control volumes}} are regions in space with constant volumetric dimensions. Here, both \textbf{\textit{mass}} and \textbf{\textit{energy}} are allowed to cross the barrier. Examples of these types of systems include nozzles or turbines. The boundaries of the \textbf{\textit{control volume}} are referred to as the \textbf{\textit{control surface}}. There are no rules to properly select a \textbf{\textit{control volume}}, but proper selection can simplify a problem's solution.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Properties of a System}
Characteristics of a \textbf{\textit{system}} are referred to as \textbf{\textit{properties}} of that \textbf{\textit{system}}. These can be mass, temperature, pressure, etc. These \textbf{\textit{properties}} can also be classified as \textbf{\textit{extensive}} or \textbf{\textit{intensive}}. The difference between the two is that \textbf{\textit{intensive properties}} are independent of the mass of the \textbf{\textit{system}}.

For our purposes, \textbf{\textit{systems}} will be treated as if the mass and matter entering and exiting the boundaries are part of a \textbf{\textit{continuum}}, or that it is homogenous. The properties do not vary with location for a given material.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Density and Specific Gravity}
\textbf{\textit{Density}} is the measure of mass per unit volume:
%
\begin{equation}
\rho = \frac{m}{\nu}
\end{equation}
%
This is the inverse of \textbf{\textit{specific volume}}. \textbf{\textit{Density}} tends to be a function of the \textbf{\textit{temperature}} and \textbf{\textit{pressure}} of a substance. Most liquids and solids are considered to be \textit{incompressible}, that is their density does not vary much with changes in \textbf{\textit{pressure}}.

\textbf{\textit{Density}} is sometimes referred to as a ratio between that of the given substance and a well-known substance, like water. This is referred to as \textbf{\textit{Specific Gravity}}:
%
\begin{equation}
\text{SG} = \frac{\rho}{\rho_{\text{H}_2 \text{O}}}
\end{equation}
%
This is a dimensionless quantity. The \textbf{\textit{specifc weight}} of a substance is given by:
%
\begin{equation}
\gamma_s = \rho g
\end{equation}
%
where $g$ is the acceleration due to gravity. This has units of $\frac{\text{N}}{\text{m}^3}$.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{State and Equilibrium}
Suppose there is a \textbf{\textit{system}} not undergoing any change. At this instant, the system's \textbf{\textit{properties}} can be either measured or calculated. The resulting values for the properties that completely describe the system's current configuration is referred to as a system's \textbf{\textit{state}}.

Thermodynamics is mostly concerned with \textbf{\textit{equilibrium states}}. This occurs when there are no internal or external forces on the system that would cause the system to change \textbf{\textit{states}} when isolated from its \textbf{\textit{surroundings}}.

\textbf{\textit{Equilibrium}} can be separated into several components, including \textbf{\textit{thermal}}, \textbf{\textit{mechanical}}, \textbf{\textit{phase}}, and \textbf{\textit{chemical}}. \textbf{\textit{Thermal equilibrium}} requires there to be no \textbf{\textit{temperature}} difference throughout the \textbf{\textit{system}}. Likewise, \textbf{\textit{mechanical equilibrium}} implies uniform pressure. \textbf{\textit{Phase equilibrium}} occurs when the mass of each phase reaches an equilibrium level and does not change. \textbf{\textit{Chemical equilibrium}} requires there to be no chemical compositional changes with time. All of these must be satisfied for a \textbf{\textit{system}} to be in an \textbf{\textit{equilibrium state}}.

\textbf{\textit{The State Postulate}} states that \textit{the state of a simple, compressible system is completely specified by two independent, intensive properties}. This type of \textbf{\textit{system}} exists if there are n electrical, magnetic, gravitational, motion, and surface tension effects. Otherwise, the \textbf{\textit{State Postulate}} does not hold.

\textbf{\textit{Intensive properties}} are \textbf{\textit{independent}} if it is possible to change one without adjusting the other. Some examples of these \textbf{\textit{intensive properties}} are \textbf{\textit{temperature}} and \textbf{\textit{specific volume}}. An example of coupled \textbf{\textit{intensive properties}} are \textbf{\textit{temperature}} and \textbf{\textit{pressure}}.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Processes and Cycles}
Any change in a \textbf{\textit{system}} that brings it from one \textbf{\textit{equilibrium state}} to another is referred to as a \textbf{\textit{process}}. The series of \textbf{\textit{states}} that the \textbf{\textit{system}} to experience during the \textbf{\textit{process}} is referred to as the \textbf{\textit{process path}}.

There are some special processes:
%
\begin{enumerate}
\item \textbf{\textit{isothermal process}} -- A \textbf{\textit{process}} where \textbf{\textit{temperature}} remains constant.\\
\item \textbf{\textit{isobaric process}} -- A \textbf{\textit{process}} where \textbf{\textit{pressure}} remains constant.\\
\textbf{\textit{isometric process}} -- A \textbf{\textit{process}} where the \textbf{\textit{specific volume}} remains constant.
\end{enumerate}
%

If a series of \textbf{\textit{processes}} result in two identical \textbf{\textit{system states}}, then the \textbf{\textit{system}} is said to have undergone a \textbf{\textit{cycle}}.

A \textbf{\textit{steady-flow process}} can be used to idealize real processes. This means that the \textbf{\textit{system mass}} is flowing through the \textbf{\textit{control volume}} steadily. This can be used to represent continuous-operation devices, such as turbines, pumps, boilers, condensers, heat exchangers, refrigeration systems. \textbf{\textit{Cyclic processes}} can also be approximated by \textbf{\textit{steady-flow processes}} if an average over time is used for \textbf{\textit{system properties}}.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Temperature and the Zeroth Law of Thermodynamics}
Recall that \texbf{\textit{thermal equilibrium}} requires \textbf{\textit{temperature}} to be uniform throughout a \textbf{\textit{system}}. The \textbf{\textit{Zeroth Law of Thermodynamics}} states that \textit{if two bodies are in thermal equilibrium with a third body, then they are also in thermal equilibrium with each other}.

This is important for forming an understanding of \textbf{\textit{temperature}}, because there are different scales for measuring \textbf{\textit{temperature}}. Thoughout this document, you may encounter the following:
%
\begin{enumerate}
\item \textbf{\textit{Celsius}} -- denoted by $\#^\circ \text{C}$\\
\item \textbf{\textit{Fahrenheit}} -- denoted by $\#^\circ \text{F}$\\
\textbf{\textit{Kelvin}} -- denoted by $\#K$\\
\textbf{\textit{Rankine}} -- denoted by $\#R$\\
\end{enumerate}
%
The following relationships can be used to convert between different scales:
%
\begin{equation}
T(\text{K}) = T(^\circ C) + 273.15
\end{equation}
%
%
\begin{equation}
T(\text{R}) = T(^\circ F) + 459.67
\end{equation}
%
%
\begin{equation}
T(^\circ F) = 1.8T(^\circ C) + 32
\end{equation}
%

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Pressure}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Pressure Measurement Devices}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Problem Solving Technique}
